\thesisTitle

### 2019-10-28 proposal/paper.tex
first draft of chosen research question & approach

\usepackage{graphicx}
\graphicspath{{images/}}

@@

Graph regression problems have for the most part only been discussed in the context of specific domain problems.
One of those domains is the prediction of chemical toxicity, typically the $\text{LD}_{50}$ or $\text{LC}_{50}$ doses which measure the smallest amount of a given chemical that is lethal for at least $50\%$ of tested subjects.
Another regression task from the domain of molecular analysis is the prediction of compound solubility.
The following approaches could serve as inspiration, baselines and/or sources of training data:
\begin{itemize}
	\item \textit{RASAR}~\cite{Luechtefeld2018} uses a logistic regression model.
		It is trained with manually chosen features that are given for training molecule graphs.
		To predict the toxicity of a new chemical, $k$-means is used to average the features of chemicals with similar molecule graphs in the training set.
		The structural similarity between molecule graphs is measured by the Tanimoto coefficient of their chemical fingerprints.
		Those fingerprints encode the presence or absence of certain meaningful chemical substructures.
		The implementation and used dataset is unfortunately not publicly available because it is part of a commercial product\footnote{ToxTrack Inc., \url{https://toxtrack.com/}}.
	\item \textit{ProTox}~\cite{Drwal2014}\cite{Banerjee2018} is another toxicity prediction approach.
		It works similarly to RASAR but leaves out the logistic regression step and instead directly aggregates the toxicity of similar chemicals via $k$-means.
		ProTox uses the freely available \textit{SuperToxic} dataset~\cite{Schmidt2009} which contains 13400 molecules with known $\text{LD}_{50}$ values.
		While the implementation is again not publicly available, the trained model can be freely queried online\footnote{ProTox-II, \url{http://tox.charite.de/protox_II/}}.
	\item The \textit{Toxicity Estimation Software Tools}~(T.E.S.T.)\footnote{\url{https://www.epa.gov/chemical-research/toxicity-estimation-software-tool-test}} is a free collection of models for toxicity prediction.
		Like RASAR and ProTox it includes similarity-based prediction models but also linear regression and decision tree models that are trained on vector representations of chemicals.
		The publicly available \textit{ChemIDplus}\footnote{\url{https://chem.nlm.nih.gov/chemidplus/}} database was used for training.
		It contains 28023 and 13541 molecules with known oral $\text{LD}_{50}$ values for rats and mice respectively.
		Additionally about 1000 molecules with known $\text{LC}_{50}$ values are available for both species.
	\item Recently first toxicity prediction approaches using ConvGNNs have been published.
		The previously mentioned AGCN~\cite{Li2018} was in fact evaluated on a discrete toxicity classification problem;
		this could serve as a starting point for a continuous graph regression method.
		\citet{Pope2018} describe how functional groups of toxic chemicals can be learned using GCNs;
		there a functional group refers to a relevant subgraph of a molecule that can be interpreted as a toxicity indicator.
	\item The previously described EGNN method~\cite{Gong2018} for multi-relational graphs was evaluated on a compound solubility prediction task.
		For this the Lipophilicity~\cite{Wu2017} and FreeSolv~\cite{Mobley2014} datasets were used.
		They contain 4200 and 642 labeled molecular graphs respectively.
\end{itemize}

\section{Goals}%
\label{sec:goals}

As mentioned in \cref{sec:motivation}, the overall goal of the thesis is to extend the LTA approach described in \cref{sec:related-work:lta} to the graph setting.
Since this goal strongly overlaps with the work on graph classification described in \cref{sec:related-work:gcr}, both branches of research should be combined.
The practical applications described in \cref{sec:related-work:tox} could be used for comparison and as a potential source of graph regression training/test datasets.
Now follows a concrete list of required and optional goals that should be achieved.

\subsection{Required Goals}%
\label{sec:goals:req}

\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Aggregation:}
		Define a class of aggregation functions that takes vertex scores as its input.
		It should fulfill the role of the uninorms and OWA functions in the existing LTA papers (see \cref{sec:related-work:lta}).
	\item \textbf{Disaggregation:}
		Describe how vertex features should be used to compute a constituent score that will be combined by the aggregation function.
	\item \textbf{Learning:}
		Combine the chosen aggregation and disaggregation methods into a joint LTA model with a suitable loss function.
	\item \textbf{Implementation:}
		A reference implementation for the designed LTA model should be developed.
	\item \textbf{Evaluation:}
		The reference implementation should be used to evaluate the model on at least one test dataset.
		Due to the relative scarcity of different graph regression datasets, the used dataset might also be synthetically generated.
		A generated dataset should be obtained using a variety of representative deterministic graph scoring measures to check which kinds of local and global structural features are learnable;
		some examples of measures that might be part of a graph scoring functions are the count of $k$-cliques in a graph, the graph diameter, the presence of cycles or various properties of a graph's degree distribution.
		The results should be compared with the results of at least one other approach for learning graph scores, e.g.\ a ConvCNN with a fixed aggregation function instead of a learned one.
\end{enumerate}

\subsection{Optional Goals}%
\label{sec:goals:opt}

If time permits, the previous goals can be extended.
For each of the previous five goal categories, optional additional ideas are listed below.
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Aggregation:}
		Multiple classes of aggregation functions could be considered and compared.
	\item \textbf{Disaggregation:}
		As described in \cref{sec:related-work:gcr} there are multiple methods to learn vertex scores.
		Depending on the chosen overall approach, a few of those methods might be suitable to be combined with an aggregation function learner.
		Those suitable combinations could be evaluated and compared.
	\item \textbf{Learning:}
		The joint model of aggregation and disaggregation function learning can most likely be optimized using various algorithms.
		Those algorithms could be compared.
		In case their runtime performance is impractical in certain situations, approximations and simplifications of the model could be considered.
	\item \textbf{Implementation:}
		The chosen approach should ideally align well with the practical constraints of implementation.
		It should be suitable for an implementation on modern GPUs and/or distributed cluster environments.
	\item \textbf{Evaluation:}
		If possible, real datasets should be used for evaluation (see \cref{sec:related-work:tox}) as an alternative or in addition to generated data.
		Aspects like runtime performance and scalability could also be evaluated.
\end{enumerate}

@@

As described in \cref{sec:motivation}, to extend LTA to graph inputs the aggregation and disaggregation problems have to be solved.
Since the various approaches to learn vertex features described in \cref{sec:related-work:gcr} can already be interpreted as solutions for the disaggregation problem, only a compatible vertex score aggregation function learning method has to be developed.
This corresponds to learning a pooling function in ConvGNNs.
The previously mentioned SortPooling and Self-Attention Pooling methods already do this in the classification setting.
A method for the regression setting has to be found.

To do so, training and test data is needed in the first step.
Therefore one or multiple of the chemical toxicity datasets mentioned in \cref{sec:related-work:tox} should be prepared for the regression pipeline.
This includes the following tasks:
\begin{enumerate*}
	\item The toxicity data has to be read from various formats (CSV, JSON or XML from a REST-API, chemical SD files, etc.),
	\item serialized SMILES strings of molecular structure need to be translated into adjacency matrices\footnote{e.g.\ using RDKit (\url{https://github.com/rdkit/rdkit}), pysmiles (\url{https://github.com/pckroon/pysmiles}) or Open Babel (\url{http://openbabel.org})} and
	\item the various units of measurement for toxicity need to be normalized.
\end{enumerate*}
In case the toxicity data turns out to be unsuitable, an artificial dataset might alternatively have to be generated as described in \cref{sec:goals:req}.
Whether this is necessary will only become apparent after the first evaluations which could show that the chosen approach is completely ill suited to the toxicity prediction task.

In the next step the development of a vertex aggregation method can begin.
For the disaggregation problem the open source reference implementation of GCNs\footnote{\url{https://github.com/tkipf/gcn}} or more generally the Graph Nets library\footnote{\url{https://github.com/deepmind/graph_nets}} for TensorFlow might be useful.
The development should start by building a simple ConvGNN with a fixed aggregation function.
Then an aggregation function learner can be added.
Since the existing GNN tools and reference implementations are mostly available in Python, the implementation part of this thesis should probably also use this language.

Lastly the developed method should be evaluated using the dataset that was chosen in the first step.
The results on the toxicity dataset should be compared with the results of selected approaches from \cref{sec:related-work:tox}.
Additionally the learned aggregation functions should be compared with na{\"\i}ve aggregation functions like $\min$, $\max$ or $\text{avg}$ since those aggregation functions are typically used in the pooling layer of current ConvGNN approaches.

@@

\begin{enumerate}
	\item Introduction
	\item Previous Work
		\begin{enumerate}[label*=\arabic*.]
			\item Learning to Aggregate
			\item Convolutional Graph Neural Networks
		\end{enumerate}
	\item Learning to Aggregate Graphs
	\begin{enumerate}[label*=\arabic*.]
		\item A Vertex Aggregation Function Learner
		\item Compatible Disaggregation Methods
	\end{enumerate}
	\item Evaluation
		\begin{enumerate}[label*=\arabic*.]
			\item Learning Characteristics of Generated Graphs
			\item Learning Chemical Toxicity
			\item Scalability
		\end{enumerate}
	\item Conclusion
	\item Literature
	\item Appendix
\end{enumerate}

@@

		\ganttgroup{Implementation}{1}{11} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Data Preparation}{1}{1} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Basic ConvGNN}{2}{4} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Aggregation Learner}{5}{8} \\
			\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellgrau}]
				{Buffer}{9}{9} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Evaluation}{10}{11} \\

		\ganttgroup{Writing}{2}{19} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Previous Work}{2}{5} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{L.t.A. Graphs}{6}{10} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Evaluation}{11}{13} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Conclusion}{14}{14} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellgrau}]
			{Buffer}{15}{17} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Corrections}{18}{19} \\

>>>>

\usepackage{graphicx}
\graphicspath{{gfx/}}

@@

Graph regression problems have for the most part only been discussed in the context of specific domain problems.
One of those domains is the prediction of chemical toxicity, typically the $\text{LD}_{50}$ or $\text{LC}_{50}$ doses which measure the smallest amount of a given chemical that is lethal for at least $50\%$ of tested subjects.
Another regression task from the domain of molecular analysis is the prediction of compound solubility.
The following approaches could serve as inspiration, baselines and/or sources of training data:
\begin{itemize}
	\item \textit{RASAR}~\cite{Luechtefeld2018} uses a logistic regression model.
		It is trained with manually chosen features that are given for training molecule graphs.
		To predict the toxicity of a new chemical, $k$-means is used to average the features of chemicals with similar molecule graphs in the training set.
		The structural similarity between molecule graphs is measured by the Tanimoto coefficient of their chemical fingerprints.
		Those fingerprints encode the presence or absence of certain meaningful chemical substructures.
		The implementation and used dataset is unfortunately not publicly available because it is part of a commercial product\footnote{ToxTrack Inc., \url{https://toxtrack.com/}}.
	\item \textit{ProTox}~\cite{Drwal2014}\cite{Banerjee2018} is another toxicity prediction approach.
		It works similarly to RASAR but leaves out the logistic regression step and instead directly aggregates the toxicity of similar chemicals via $k$-means.
		ProTox uses the freely available \textit{SuperToxic} dataset~\cite{Schmidt2009} which contains 13400 molecules with known $\text{LD}_{50}$ values.
		While the implementation is again not publicly available, the trained model can be freely queried online\footnote{ProTox-II, \url{http://tox.charite.de/protox_II/}}.
	\item The \textit{Toxicity Estimation Software Tools}~(T.E.S.T.)\footnote{\url{https://www.epa.gov/chemical-research/toxicity-estimation-software-tool-test}} is a free collection of models for toxicity prediction.
		Like RASAR and ProTox it includes similarity-based prediction models but also linear regression and decision tree models that are trained on vector representations of chemicals.
		The publicly available \textit{ChemIDplus}\footnote{\url{https://chem.nlm.nih.gov/chemidplus/}} database was used for training.
		It contains 28023 and 13541 molecules with known oral $\text{LD}_{50}$ values for rats and mice respectively.
		Additionally about 1000 molecules with known $\text{LC}_{50}$ values are available for both species.
	\item Recently first toxicity prediction approaches using ConvGNNs have been published.
		The previously mentioned AGCN~\cite{Li2018} was in fact evaluated on a discrete toxicity classification problem;
		this could serve as a starting point for a continuous graph regression method.
		\citet{Pope2018} describe how functional groups of toxic chemicals can be learned using GCNs;
		there a functional group refers to a relevant subgraph of a molecule that can be interpreted as a toxicity indicator.
	\item The previously described EGNN method~\cite{Gong2018} for multi-relational graphs was evaluated on a compound solubility prediction task.
		For this the Lipophil\-icity~\cite{Wu2017} and FreeSolv~\cite{Mobley2014} datasets were used.
		They contain 4200 and 642 labeled molecular graphs respectively.
\end{itemize}

\section{Research Questions}%
\label{sec:question}

When looking only at the input and output of the described LTA and graph methods, they share certain similarities.
Both take a variable size and permutation invariant set of constituents/vertices as input and compute an aggregate class or score from it.
There is however a difference between LTA and the described graph methods in the way aggregate scores are computed.

LTA approaches are based on the assumption that a composite score can be modeled as an aggregate of constituent scores where constituent and composite scores are required to be from the same domain, e.g.\ $\mathcal{Y} = [0, 1]$.
One consequence of this assumption is that an aggregate score is explainable by looking at the constituents and their associated scores.

A graph classification or regression method on the other hand does not necessarily output explainable classes or scores.
ConvGNNs for example perform multiple neighborhood feature aggregations around each vertex and then combine the aggregated vertex features into a final output using either a fixed aggregation function (e.g.\ $\min$, $\max$ or $\text{avg}$) or a standard predictor that gets a fixed-size top-$k$ ranking of the vertex features.
A given aggregate score is generally not explained by the presence or absence of certain constituents.

Intuitively the difference between LTA and current graph methods can be formulated as \textit{first evaluate, then aggregate} versus \textit{first aggregate, then evaluate}, see \cref{fig:gnnvslta}.
This general difference is however not a strict separator of LTA and current graph methods.
As described by \citet{Pope2018} for example, relevant subgraphs with a high influence on the output of a ConvGNN can be identified and interpreted similarly to LTA constituents.
The line between both approaches is therefore blurry.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\linewidth]{GNNvsLTA.pdf}
	\caption{Intuition for the difference between ConvGNNs and LTA methods.}\label{fig:gnnvslta}
\end{figure}

To address this, the thesis should try to answer three central questions:
\begin{enumerate}
	\item \textit{What are the essential characteristics of a Learning to Aggregate method?}
		Those characteristics should formally capture the differences and similarities between LTA and convolutional or kernel based graph methods.
		Currently there is no comprehensive formulation of the relation between both fields of research;
		this should be addressed by the first question.
	\item \textit{How could a Learning to Aggregate method, that has the previously defined characteristics, look like for graph inputs?}
		Currently LTA has only been applied to sets of constituents and not yet to graph structured data;
		this should be addressed by the second question.
	\item \textit{What are the assumptions of graph LTA methods and under which circumstances are they applicable?}
		This aims to clarify in which kinds of scenarios the use of LTA is justified and expected to improve results compared to more general methods.
\end{enumerate}

@@

To answer the three research questions, the following approach will be taken.
First the notion of composites and constituents should be generalized based on the definitions found in the previous works on LTA~\cite{Melnikov2016}\cite{Melnikov2019}, so that they are also applicable to graph inputs.
Next the notion of an aggregation function has to be defined and distinguished from the more general concept of an evaluation function (compare \cref{fig:gnnvslta}).

The resulting general definition of LTA methods should then be used for a comparison with a representative selection of existing graph methods (see \cref{sec:related-work:gcr}).
It should be stated how both fields of research are related and under which circumstances a graph method can be interpreted as an LTA method.
When formulating a graph method from the LTA perspective, it must be explicitly stated how the constituents and the aggregation function looks like.
Depending on the graph method a constituent could consist of a single vertex but could also represent a certain subgraph or characteristic graph pattern.

The comparison between LTA and existing graph methods should serve as an overview of the state-of-the-art in graph classification and regression from the perspective of LTA.\@
Based on this comparison, a specific graph LTA method or class of graph LTA methods should be developed.
Depending on the interpretability of existing graph methods as LTA methods, the developed method will potentially be based on those existing methods.

Finally the proposed graph LTA method should be evaluated.
For this synthetically generated data or some of the datasets listed in \cref{sec:related-work:tox} could be used.
If time permits, a comparison between the developed graph LTA method and other graph methods should be conducted to empirically demonstrate the potential advantages and disadvantages of the LTA assumptions.

@@

\begin{enumerate}
	\item Introduction
	\item Previous Work
		\begin{enumerate}[label*=\arabic*.]
			\item Learning to Aggregate
			\item Convolutional Graph Neural Networks
		\end{enumerate}
	\item Learning to Aggregate on Graphs
	\begin{enumerate}[label*=\arabic*.]
		\item Characteristics of LTA methods
		\item An LTA perspective on GNNs
		\item A Graph LTA Method
	\end{enumerate}
	\item Evaluation
	\item Conclusion
	\item Literature
	\item Appendix
\end{enumerate}

@@

		\ganttgroup{Implementation}{2}{12} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Data Preparation}{2}{2} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Graph LTA}{3}{8} \\
			\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellgrau}]
				{Buffer}{9}{10} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Evaluation}{11}{12} \\

		\ganttgroup{Writing}{1}{19} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Previous Work}{1}{4} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Graph LTA}{5}{10} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Evaluation}{11}{13} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Conclusion}{14}{14} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellgrau}]
			{Buffer}{15}{17} \\
		\ganttbar[bar label font = \sffamily, bar/.append style = {fill = hellblau}]
			{Corrections}{18}{19} \\
